# MLM, Longitudinal: Hox ch 5 - student GPA



```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("sarbearschwartz/apaSupp") # updated 9/17/2025
```

Load/activate these packages

```{r, message=FALSE, error=FALSE, warning=FALSE, results='asis'}
library(haven)     
library(tidyverse)    
library(flextable)
library(apaSupp)      # Not on CRAN, on GitHub (see above)
library(psych)
library(VIM)         
library(naniar) 
library(lme4)        
library(lmerTest)
library(optimx)
library(performance)
library(interactions)
library(effects)
library(emmeans)
library(ggResidpanel)
library(HLMdiag)
library(sjPlot)
library(sjstats)     
```




## Background


```{block type='rmdlink', echo=TRUE}
The text **"Multilevel Analysis: Techniques and Applications, Third Edition"** [@hox2017] has a companion [website](https://multilevel-analysis.sites.uu.nl/) which includes links to all the data files used throughout the book (housed on the [book's GitHub repository](https://github.com/MultiLevelAnalysis)).  
```


The following example is used through out @hox2017's chapater 5.

The GPA for 200 college students were followed for 6 consecutive semesters (simulated).  Job status was also measured as number of hours worked for the same size occations.  Time-invariant covariates are the student's gender and high school GPA.  The variable `admitted` will not be used.  


```{r}
df_raw <- haven::read_sav("data/gpa2long.sav") %>% 
  haven::as_factor() %>%             # retain the labels from SPSS --> factor
  haven::zap_label()
```


```{r}
tibble::glimpse(df_raw) 
```

```{r}
df_raw %>% 
  dplyr::select(occas, job) %>% 
  table(useNA = "always")
```



```{r}
df_gpa_long <- df_raw %>% 
  dplyr::mutate(student = factor(student)) %>% 
  dplyr::mutate(sem = case_when(occas == "year 1 semester 1" ~ 1,
                                occas == "year 1 semester 2" ~ 2,
                                occas == "year 2 semester 1" ~ 3,
                                occas == "year 2 semester 2" ~ 4,
                                occas == "year 3 semester 1" ~ 5,
                                occas == "year 3 semester 2" ~ 6)) %>% 
  dplyr::mutate(job = fct_drop(job)) %>% 
  dplyr::mutate(hrs = case_when(job == "no job" ~ 0,
                                job == "1 hour" ~ 1,
                                job == "2 hours" ~ 2,
                                job == "3 hours" ~ 3,
                                job == "4 or more hours" ~ 4)) %>%  
  dplyr::select(student, sex, highgpa, sem, job, hrs, gpa) %>% 
  dplyr::arrange(student, sem)
```

```{r}
tibble::glimpse(df_gpa_long)
```


```{r}
df_gpa_long %>% 
  psych::headTail(top = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Data in Long Format")
```


```{r}
df_gpa_wide <- df_gpa_long %>% 
  tidyr::pivot_wider(names_from = sem,
                     values_from = c(job, hrs, gpa),
                     names_sep = "_")
```


```{r}
tibble::glimpse(df_gpa_wide)
```


```{r}
df_gpa_wide %>% 
  dplyr::select(sex,
                "High School GPA" = highgpa, 
                "Initial College GPA" = gpa_1, 
                "Initial Job" = job_1, 
                "Initial Hrs" = hrs_1) %>% 
  apaSupp::tab1(split = "sex",
                caption = "Demographic Summary by Sex")
```



```{r}
df_gpa_wide %>% 
  dplyr::select(sex, gpa_1, gpa_2, gpa_3, gpa_4, gpa_5, gpa_6) %>% 
  apaSupp::tab1(split = "sex",
                caption = "Hox Table 5.2 (page 77) GPA means at six occations, for male and female students")
```

## MLM

### Null Models and ICC



```{r}
fit_lmer_0_re <- lmerTest::lmer(gpa ~ 1 + (1|student),
                                data = df_gpa_long,
                                REML = TRUE)
```

```{r}
performance::icc(fit_lmer_0_re)
```


Over a third of the variance in the 6 GPA measures is variance between individuals, and about two-thirds is variance within individuals across time, $\rho = .369$.


```{r}
fit_lmer_1_re <- lmerTest::lmer(gpa ~ I(sem - 1) + (1|student),
                                data = df_gpa_long,
                                REML = TRUE)
```

```{r}
performance::icc(fit_lmer_1_re)
```

After accounting for the linear change in GPA over semesters, about half of the remaining variance in GPA scores is attributable to person-to-person difference.  

> Another way to say it, is about half of the variance in initial GPAs is due to student-to-student differences.

```{r}
apaSupp::tab_lmers(list("M0: Null w/o Time" = fit_lmer_0_re,
                        "M1: Null w/Time"   = fit_lmer_1_re),
                   caption = "Null Models")
```



### Fixed Effects


```{r}
fit_lmer_0_ml <- lmerTest::lmer(gpa ~ 1 + (1|student),
                                data = df_gpa_long,
                                REML = FALSE)

fit_lmer_1_ml <- lmerTest::lmer(gpa ~ I(sem - 1) + (1|student),
                                data = df_gpa_long,
                                REML = FALSE)


fit_lmer_2_ml <- lmerTest::lmer(gpa ~ I(sem - 1) + hrs + (1|student),
                                data = df_gpa_long,
                                REML = FALSE)


fit_lmer_3_ml <- lmerTest::lmer(gpa ~ I(sem - 1) + hrs + highgpa + sex + (1|student),
                                data = df_gpa_long,
                                REML = FALSE)
```


```{r}
apaSupp::tab_lmers(list("M2: Occ" = fit_lmer_1_ml, 
                        "M3: Job" = fit_lmer_2_ml, 
                        "M4: GPA, Sex" = fit_lmer_3_ml),
                   var_labels = c("(Intercept)" = "(Intercept)",
                                  "I(sem - 1)" = "Semester",
                                  "hrs" = "Hours Working",
                                  "highgpa" = "High School GPA",
                                  "sex" = "Sex",
                                  "male" = "Male",
                                  "female" = "Female"),
                   narrow = TRUE,
                   caption = "Hox Table 5.3 (page 78) Results of Multilevel Anlaysis of GPA, Fixed Effects",
                   general_note = "Intercept refers to population mean for a Male who is not working during their first semester")
```


```{r}
tab_lmer_fits(list("M0: Null w/o Time" = fit_lmer_0_re,
                   "M1: Null w/Time"   = fit_lmer_1_re, 
                   "M2: Occ"           = fit_lmer_1_ml, 
                   "M3: Job"           = fit_lmer_2_ml, 
                   "M4: GPA, Sex"      = fit_lmer_3_ml))
```





```{r}
anova(fit_lmer_0_ml, fit_lmer_1_ml, fit_lmer_2_ml, fit_lmer_3_ml)
```


### Variance Explained by linear TIME at Level ONE

```{r}
lme4::VarCorr(fit_lmer_0_ml)  # baseline
```

```{r}
lme4::VarCorr(fit_lmer_1_ml)  # model to compare
```


```{r}
lme4::VarCorr(fit_lmer_0_ml) %>%  # baseline
  print(comp = c("Variance", "Std.Dev"),
        digits = 3)
```

```{r}
lme4::VarCorr(fit_lmer_1_ml) %>%  # model to compare
  print(comp = c("Variance", "Std.Dev"),
        digits = 3)
```


```{r}
performance::r2(fit_lmer_1_ml)
```


```{r}
performance::r2(fit_lmer_1_ml, by_group = TRUE)
```



#### Raudenbush and Bryk

* Explained variance is a proportion of first-level variance only
* A good option when the multilevel sampling process is is close to two-stage simple random sampling

```{block type='genericEq', echo=TRUE}
**Raudenbush and Bryk Approximate Formula - Level 1 ** *approximate*
$$
approx \;R^2_1 = \frac{\sigma^2_{e-BL} - \sigma^2_{e-MC}}
{\sigma^2_{e-BL} }
\tag{Hox 4.8}
$$
```



```{r}
(0.098 - 0.058) / 0.098
```


#### Snijders and Bosker


```{block type='genericEq', echo=TRUE}
**Snijders and Bosker Formula - Level 1 ** 

Random Intercepts Models Only, *address potential negative $R^2$ issue*
$$
R^2_1 = 1 - \frac{\sigma^2_{e-MC} + \sigma^2_{u0-MC}}
{\sigma^2_{e-BL} + \sigma^2_{u0-BL}}
$$
```




```{r}
1 - (0.058 + 0.063)/(0.098 + 0.057)
```


### Variance Explained by linear TIME at Level TWO

#### Raudenbush and Bryk

```{block type='genericEq', echo=TRUE}
**Raudenbush and Bryk Approximate Formula - Level 2 ** 
$$
approx \; R^2_s = \frac{\sigma^2_{u0-BL} - \sigma^2_{u0-MC}}
{\sigma^2_{u0-BL} }
\tag{Hox 4.9}
$$
```





```{r}
(0.057 - 0.063)/ 0.057
```

YIKES!  Negative Variance explained!


#### Snijders and Bosker


```{block type='genericEq', echo=TRUE}
**Snijders and Bosker Formula Extended - Level 2 ** 
$$
R^2_2 = 1 - \frac{\frac{\sigma^2_{e-MC}}{B} + \sigma^2_{u0-MC}}
{\frac{\sigma^2_{e-BL}}{B} + \sigma^2_{u0-BL}}
$$

$B$ is the average size of the Level 2 units.  Technically, you should use the *harmonic mean*, but unless the clusters differ greatly in size, it doesn't make a huge difference.
```


```{r}
1 - (0.058/6 + 0.063) / (0.098/6 + 0.057)
```

Reason: The intercept only model overestimates the variance at the occasion level and underestimates the variance at the subject level (see chapter 4 of @hox2017)




```{r}

```



### Random Effects



```{r}
fit_lmer_3_re <- lmerTest::lmer(gpa ~ I(sem-1) + hrs + highgpa + sex + 
                                  (1|student),
                                data = df_gpa_long,
                                REML = TRUE)


fit_lmer_4_re <- lmerTest::lmer(gpa ~ I(sem-1) + hrs + highgpa + sex +
                                  (I(sem-1)|student),
                                data = df_gpa_long,
                                REML = TRUE,
                                control = lmerControl(optimizer ="Nelder_Mead"))
```


```{r}
anova(fit_lmer_3_re, fit_lmer_4_re, refit = FALSE)
```

### Cross-Level Interaction

```{r}
fit_lmer_4_ml <- lmerTest::lmer(gpa ~ I(sem-1) + 
                                  hrs + highgpa + sex + 
                                  (I(sem-1)|student),
                                data = df_gpa_long,
                                REML = FALSE)

fit_lmer_4_re <- lmerTest::lmer(gpa ~ I(sem-1) + 
                                  hrs + highgpa + sex + 
                                  (I(sem-1)|student),
                                data = df_gpa_long,
                                REML = TRUE,
                                control = lmerControl(optimizer ="Nelder_Mead"))
```


```{r}
fit_lmer_5_ml <- lmerTest::lmer(gpa ~ I(sem-1) + 
                                  hrs + highgpa + sex + 
                                  I(sem-1):sex + 
                                  (I(sem-1)|student),
                                data = df_gpa_long,
                                REML = FALSE)

fit_lmer_5_re <- lmerTest::lmer(gpa ~ I(sem-1) + 
                                  hrs + highgpa + sex + 
                                  I(sem-1):sex + 
                                  (I(sem-1)|student),
                                data = df_gpa_long,
                                REML = TRUE)
```


```{r}
apaSupp::tab_lmers(list("M5: Occ Rand"   = fit_lmer_4_re,
                        "M6: Xlevel Int" = fit_lmer_5_re),
                   var_labels = c("(Intercept)" = "(Intercept)",
                                  "I(sem - 1)" = "Semester",
                                  "hrs" = "Hours Working",
                                  "sexfemale" = "Sex: Female vs. Male",
                                  "highgpa" = "High School GPA",
                                  "I(sem - 1):sexfemale" = "Semester X Sex"),
                   caption = "Hox Table 5.4 (page 80) Results of Multilevel Anlaysis of GPA, Varying Effects for Occation",
                   general_note = "Intercept refers to population mean for a Male who is not working during their first semester and had a high school GPA = 0.",
                   d = 3)
```


```{r}
anova(fit_lmer_4_ml, fit_lmer_5_ml)
```


### Visualize the Model

```{r}
apaSupp::tab_lmer(fit_lmer_5_re,
                  d = 3)
```



#### Using `interactions::interact_plot()`


```{r, fig.cap="Hox Figure 5.4 (page 82) Multilevel model (M6) comapring linear increase in GPA over semester, but student's sex.", warning=FALSE, message=FALSE}
interactions::interact_plot(model = fit_lmer_5_re,
                            pred = sem,
                            modx = sex,
                            interval = TRUE)
```



```{r, warning=FALSE, message=FALSE}
interactions::interact_plot(model = fit_lmer_5_re,
                            pred = sem,
                            modx = sex,
                            mod2 = hrs,
                            mod2.values = c(1, 2, 3),
                            interval = TRUE)
```

```{r, warning=FALSE, message=FALSE}
interactions::interact_plot(model = fit_lmer_5_re,
                            pred = sem,
                            modx = sex,
                            mod2 = highgpa,
                            mod2.values = c(2, 3, 4),
                            interval = TRUE)
```

#### Using `emmeans::emmeans()` & `ggplot`

```{r}
fit_lmer_5_re %>% 
  emmeans::emmeans(~sem*sex,
                   at = list(sem = 1:6))
```


```{r}
fit_lmer_5_re %>% 
  emmeans::emmeans(~sem*sex,
                   at = list(sem = 1:6)) %>% 
  data.frame() %>% 
  dplyr::mutate(sex = sex %>% 
                  forcats::fct_recode("Male" = "male",
                                      "Female" = "female") %>% 
                  forcats::fct_rev()) %>% 
  ggplot(aes(x = sem,
             y = emmean,
             group = sex)) +
  geom_ribbon(aes(ymin = emmean - SE,
                  ymax = emmean + SE,
                  fill = sex),
              alpha = .2) +
  geom_point(aes(shape = sex,
                 color = sex),
             size = 3) +
  geom_line(aes(linetype = sex,
                color = sex)) +
  theme_bw() +
  labs(x = "Semester",
       y = "Estimated Marginal Mean\nCollege GPA",
       fill = NULL,
       color = NULL,
       shape = NULL,
       linetype = NULL) +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(breaks = seq(from = 0, to = 4, by = .2)) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  theme(legend.position = "inside",
        legend.position.inside = c(0, 1),
        legend.justification = c(-.1, 1.1),
        legend.key.width = unit(1.5, "cm"),
        legend.background = element_rect(color = "black"))
```





### Effect Sizes

#### Standardized Parameters

```{r}
effectsize::standardize_parameters(fit_lmer_5_re)
```


#### R-squared type measures

```{r}
performance::r2(fit_lmer_5_re)
```

```{r}
performance::r2(fit_lmer_5_re, by_group = TRUE)
```





#### Pairwise Differences in Means

```{r}
fit_lmer_5_re %>% 
  emmeans::emmeans(~sex | sem,
                   at = list(sem = 1:6))
```

```{r}
fit_lmer_5_re %>% 
  emmeans::emmeans(~sex | sem,
                   at = list(sem = 1:6)) %>% 
  pairs()
```



```{r}
fit_lmer_5_re %>% 
  emmeans::emmeans(~sex | sem,
                   at = list(sem = 1:6)) %>% 
  pairs(adjust = "none") %>% 
  data.frame() %>% 
  dplyr::mutate(p_Unadj = apaSupp::p_num(p.value)) %>% 
  dplyr::mutate(p_Adj = apaSupp::p_num(p.adjust(p.value, method = "fdr"))) %>% 
  dplyr::mutate(SMD = estimate/apaSupp::lmer_sd(fit_lmer_5_re)) %>% 
  dplyr::select("Semester" = sem,
                "EMM Difference_Est" = estimate, 
                "EMM Difference_(SE)" = SE,
                SMD,
                p_Unadj,
                p_Adj) %>% 
    flextable::flextable() %>% 
  flextable::separate_header() %>% 
  apaSupp::theme_apa(caption = "Sex Differences in Estimated Marginal Mean Collegate GPA for each Semester",
                     general_note = "EMM = estimated marginal means. SMD = standardized mean difference. P-values given both unadjusted (Unadj) and adjusted (Adj) via the method of Benjamini, Hochberg, and Yekutieli to control the false discovery rate (FDR)") %>% 
  flextable::hline(part = "header", i = 1)
```


#### Simple Slopes

```{r}
fit_lmer_5_re %>% 
  emmeans::emtrends( ~ sex, var = "sem")
```




### Significance

#### Fixed Effects

> The Likelyhood Ratio Test (Deviance Difference Test) is best for establishing significance of fixed effects.

Wald-tests

```{r}
summary(fit_lmer_5_re)
```

F-test with Satterthwaite adjusted degrees of freedom 

```{r}
anova(fit_lmer_5_re)
```



#### Random Effects

Likelyhood Ratio Tests (Deviance Difference Test), by single term deletion

```{r}
lmerTest::ranova(fit_lmer_5_re)
```



