# MLM, Longitudinal: RCT - Exercise and Diet



```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("sarbearschwartz/apaSupp") # updated 9/17/2025
```

Load/activate these packages

```{r, message=FALSE, error=FALSE, warning=FALSE, results='asis'}
library(tidyverse)    
library(flextable)
library(apaSupp)      # Not on CRAN, on GitHub (see above)
library(psych)
library(VIM)         
library(naniar) 
library(lme4)        
library(lmerTest)
library(optimx)
library(performance)
library(interactions)
library(effects)
library(emmeans)
library(ggResidpanel)
library(HLMdiag)
library(sjPlot)
library(sjstats)     
```


## The dataset

This comes from a **Randomized Controled Trial**.

```{r}
df_ex_raw <- read.table("data/exercise_diet.txt",
                        header = TRUE, 
                        sep = ",")
```

```{r}
tibble::glimpse(df_ex_raw)
```



```{r}
df_ex_long <- df_ex_raw %>% 
  dplyr::mutate(id = id %>% factor) %>% 
  dplyr::mutate(exertype = exertype %>% 
                  factor(levels = 1:3,
                         labels = c("At Rest",
                                    "Leisurely Walking",
                                    "Moderate Running"))) %>% 
  dplyr::mutate(diet = diet %>% 
                  factor(levels = 1:2,
                         labels = c("low-fat",
                                    "non-fat"))) %>% 
  dplyr::mutate(time_min = time / 60)
```




```{r}
df_ex_long %>% 
  psych::headTail(top = 10, bottom = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Raw Data")
```

## Exploratory Data Analysis

### Participant Summary

In this experiment, both exercise (`exertype`) and diet (`diet`) were randomized at the subject level to create a 2x3 = 6 combinations each with exactly 5 participants.

```{r}
df_ex_long %>% 
  dplyr::filter(time == 0) %>% 
  dplyr::select(exertype,
                "Diet, randomized" = diet) %>% 
  apaSupp::tab1(split = "exertype",
                total = FALSE,
                test = FALSE,
                caption = "Participants")
```

### Baseline Summary



```{r}
df_ex_long %>% 
  dplyr::filter(time == 0) %>% 
  dplyr::group_by(exertype, diet) %>% 
  dplyr::summarise(mean = mean(pulse)) %>% 
  dplyr::ungroup() %>% 
  tidyr::pivot_wider(names_from = diet,
                     values_from = mean) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Baseline Pulse, Group Means")

```

### Raw Trajectories - Person Profile Plot

#### Connect the dots

```{r}
df_ex_long %>% 
  ggplot(aes(x = time_min,
             y = pulse)) +
  geom_point() +
  geom_line(aes(group = id)) +
  facet_grid(diet ~ exertype) +
  theme_bw()
```

#### Loess - Moving Average Smoothers

```{r}
df_ex_long %>% 
  ggplot(aes(x = time_min,
             y = pulse,
             color = diet)) +
  geom_line(aes(group = id)) +
  facet_grid(~ exertype) +
  theme_bw() +
  geom_smooth(method = "loess",
              se = FALSE,
              size = 2,
              span = 5) +
  theme(legend.position = c(0.08, 0.85),
        legend.background = element_rect(color = "black")) +
  labs(title = "Raw Pulse Trajectories",
       subtitle = "By Exercise and Diet Groupings",
       x = "Time (Minutes Post-Baseline)",
       y = "Pulse (Beats per Minute)",
       color = "Diet Plan")
```




## Multilevel Modeling

### Null Model

```{r}
fit_lmer_0re <- lmerTest::lmer(pulse ~ 1 + (1 | id),
                               data = df_ex_long)
```


```{r}
apaSupp::tab_lmer(fit_lmer_0re)
```

### ICC & R-squared

```{r}
performance::icc(fit_lmer_0re)
```


```{r}
performance::r2(fit_lmer_0re)
```


### Add fixed effects: level specific

#### Fit nested models

```{r}
# Null Model (random intercept only)
fit_lmer_0ml <- lmerTest::lmer(pulse ~ 1 + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)

# Add quadratic time
fit_lmer_1ml <- lmerTest::lmer(pulse ~ time_min + I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)

# Add main effects for 2 interventions (person-specific, i.e. level-2 factors)
fit_lmer_2ml <- lmerTest::lmer(pulse ~ diet + exertype + time_min + I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)

# Add interaction between level-2 factors
fit_lmer_3ml <- lmerTest::lmer(pulse ~ diet*exertype + time_min + I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)

# Add exercise interacting with [time & time-squared]
fit_lmer_4ml <- lmerTest::lmer(pulse ~ diet*exertype + exertype*time_min + exertype*I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)

# Add diet interacting with [time & time-squared]
fit_lmer_5ml <- lmerTest::lmer(pulse ~ diet*exertype*time_min + diet*exertype*I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = FALSE)
```


```{r}
apaSupp::tab_lmers(list("M0" = fit_lmer_0re,
                        "M1" = fit_lmer_1ml, 
                        "M2" = fit_lmer_2ml),
                   narrow = TRUE)
```
```{r}
apaSupp::tab_lmers(list("M3" = fit_lmer_3ml,
                        "M4" = fit_lmer_4ml, 
                        "M5" = fit_lmer_5ml),
                   narrow = TRUE)
```


#### Evaluate Model Fit, i.e. variable significance

```{r}
anova(fit_lmer_1ml, 
      fit_lmer_2ml, 
      fit_lmer_3ml, 
      fit_lmer_4ml, 
      fit_lmer_5ml)
```

```{r}
apaSupp::tab_lmer_fits(list("M1" = fit_lmer_1ml, 
                            "M2" = fit_lmer_2ml, 
                            "M3" = fit_lmer_3ml, 
                            "M4" = fit_lmer_4ml, 
                            "M5" = fit_lmer_5ml))
```



### Final Model

Refit via REML

```{r}
fit_lmer_5re <- lmerTest::lmer(pulse ~ diet*exertype*time_min + 
                                 diet*exertype*I(time_min^2) + (1 | id),
                               data = df_ex_long,
                               REML = TRUE)
```


#### Visualize

```{r}
sjPlot::plot_model(fit_lmer_5re,
                   type = "pred",
                   terms = c("time_min", "diet", "exertype"))
```

```{r}
interactions::interact_plot(fit_lmer_5re,
                            pred = time_min,
                            modx = diet,
                            mod2 = exertype,
                            interval = TRUE)
```




```{r}
effects::Effect(focal.predictors = c("diet", "exertype", "time_min"),
                mod = fit_lmer_5re) %>% 
  data.frame %>% 
  ggplot(aes(x = time_min,
             y = fit,
             fill = diet,
             color = diet)) +
  geom_line(size = 1.5) +
  theme_bw() +
  facet_grid(~ exertype) +
  theme(legend.position = c(0.08, 0.85),
        legend.background = element_rect(color = "black")) +
  labs(title = "Raw Pulse Trajectories",
       subtitle = "By Exercise and Diet Groupings",
       x = "Time (Minutes Post-Baseline)",
       y = "Estimated Marginal Mean\nPulse (Beats per Minute)",
       fill = "Diet Plan",
       color = "Diet Plan")
```


```{r}
effects::Effect(focal.predictors = c("diet", "exertype", "time_min"),
                mod = fit_lmer_5re,
                xlevels = list("time_min" = seq(from = 0, 
                                                to   = 12, 
                                                by   = 0.5))) %>% 
  data.frame %>% 
  dplyr::mutate(diet = fct_rev(diet)) %>%  # reverse the order of the levels
  ggplot(aes(x = time_min,
             y = fit)) +
  geom_ribbon(aes(ymin = fit - se,
                  ymax = fit + se,
                  fill = diet),
              alpha = 0.3) +
  geom_line(aes(linetype = diet),
            size = 1) +
  theme_bw() +
  facet_grid(~ exertype) +
  theme(legend.position = c(0.12, 0.85),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  labs(title = "Raw Pulse Trajectories",
       subtitle = "By Randomized Exercise and Diet Intervention",
       x = "Time (Minutes Post-Baseline)",
       y = "Estimated Marginal Mean\nPulse (Beats per Minute)",
       fill = "Diet Plan",
       color = "Diet Plan",
       linetype = "Diet Plan") +
  scale_fill_manual(values = c("black", "gray50")) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_x_continuous(breaks = seq(from = 0, to = 14, by = 5))
```


```{r}
fit_lmer_5re %>% 
  emmeans::emmeans(~ diet*exertype*time_min,
                   at = list("time_min" = seq(from = 0, 
                                                to   = 12, 
                                                by   = 0.5))) %>% 
  data.frame %>% 
  dplyr::mutate(diet = fct_rev(diet)) %>%  # reverse the order of the levels
  ggplot(aes(x = time_min,
             y = emmean)) +
  geom_ribbon(aes(ymin = emmean - SE,
                  ymax = emmean + SE,
                  fill = diet),
              alpha = 0.3) +
  geom_line(aes(linetype = diet),
            size = 1) +
  theme_bw() +
  facet_grid(~ exertype) +
  theme(legend.position = c(0.12, 0.85),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(2, "cm")) +
  labs(title = "Raw Pulse Trajectories",
       subtitle = "By Randomized Exercise and Diet Intervention",
       x = "Time (Minutes Post-Baseline)",
       y = "Estimated Marginal Mean\nPulse (Beats per Minute)",
       fill = "Diet Plan",
       color = "Diet Plan",
       linetype = "Diet Plan") +
  scale_fill_manual(values = c("black", "gray50")) +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_x_continuous(breaks = seq(from = 0, to = 14, by = 5))
```


#### Pairwise Differences in Means



```{r}
fit_lmer_5re %>% 
  emmeans::emmeans(~diet|exertype*time_min,
                   at = list(time_min = c(0, 5, 10))) %>% 
  pairs(adjust = "none") %>% 
  data.frame() %>% 
  dplyr::mutate(p_Unadj = apaSupp::p_num(p.value)) %>% 
  dplyr::mutate(p_Adj = apaSupp::p_num(p.adjust(p.value, method = "fdr"))) %>% 
  dplyr::mutate(SMD = estimate/apaSupp::lmer_sd(fit_lmer_5re)) %>% 
  dplyr::arrange(exertype, time_min) %>% 
  dplyr::select("Exercise" = exertype,
                "Time, min" = time_min,
                "EMM Difference_Est" = estimate, 
                "EMM Difference_(SE)" = SE,
                SMD,
                p_Unadj,
                p_Adj) %>% 
  flextable::as_grouped_data(groups = "Exercise") %>% 
  flextable::flextable() %>% 
  flextable::separate_header() %>% 
  apaSupp::theme_apa(caption = "Diet Differences in Estimated Marginal Mean Pulse for each Exercise Program",
                     general_note = "EMM = estimated marginal means. SMD = standardized mean difference. P-values given both unadjusted (Unadj) and adjusted (Adj) via the method of Benjamini, Hochberg, and Yekutieli to control the false discovery rate (FDR)") %>% 
  flextable::hline(part = "header", i = 1)
```

