# GEE, Continuous Outcome: Beat the Blues

## PREPARATION 

### Load Packages


```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("sarbearschwartz/apaSupp") # updated 10/14/2025
```

Load/activate these packages

```{r, message=FALSE, error=FALSE, warning=FALSE, results='asis'}
library(tidyverse)    
library(flextable)
library(apaSupp)      # Not on CRAN, on GitHub (see above)
library(psych) 
library(lme4)         
library(lmerTest)
library(corrplot)     
library(gee)          
library(geepack)       
library(MuMIn)  
library(interactions)
library(performance)      

library(HSAUR)        # package with the dataset
```




### Background


> This dataset was used as an example in Chapter 11 of "A Handbook of Statistical Analysis using R" by Brian S. Everitt and Torsten Hothorn.  The authors include this data set in their **HSAUR** package on `CRAN`. 


**The Background**


"These data resulted from a clinical trial of an interactive multimedia program called 'Beat the Blues' designed to deliver cognitive behavioral therapy (CBT) to depressed patients via a computer terminal.  Full details are given [here: Proudfoot et. al (2003)](https://www.cambridge.org/core/journals/psychological-medicine/article/computerized-interactive-multimedia-cognitivebehavioural-program-for-anxiety-and-depression-in-general-practice/6FF75F9283DF2209226D8910921F5307), but in essence Beat the Blues is an interactive program using multimedia techniques, in particular video vignettes."

"In a randomized controlled trial of the program, patients with depression recruited in primary care were randomized to either the Beating the Blues program or to "Treatment as Usual" (TAU). Patients randomized to the BEat the Blues also received pharmacology and/or general practice (GP) support and practical/social help,offered as part of treatment as usual, with the exception of any face-to-face counseling or psychological intervention.  Patients allocated to TAU received whatever treatment their GP prescribed.  The latter included, besides any medication, discussion of problems with GP, provisions of practical/social help, referral to a counselor, referral to a practice nurse, referral to mental health professionals, or further physical examination.

**The Research Question**

Net of other factors (use of antidepressants and length of the current episode), does the Beat-the-Blues program results in better depression trajectories over treatment as usual?


**The Data**

The variables are as follows:   

* `drug` did the patient take anti-depressant drugs (No or Yes)    
* `length`  the length of the current episode of depression, a factor with levels:    
+ "<6m" *less than six months*     
+ ">6m" *more than six months*   
* `treatment` treatment group, a factor with levels:   
+ "TAU" *treatment as usual*     
+ "BtheB" *Beat the Blues*    
* `bdi.pre` Beck Depression Inventory II, before treatment    
* `bdi.2m` Beck Depression Inventory II, after 2 months    
* `bdi.4m` Beck Depression Inventory II, after 4 months    
* `bdi.6m` Beck Depression Inventory II, after 6 months    
* `bdi.8m` Beck Depression Inventory II, after 8 months    


### Read in the data

```{r}
data(BtheB, package = "HSAUR")

BtheB %>% 
  psych::headTail()
```


#### Wide Format

Tidy up the dataset

**WIDE format**
* One line per person
* Good for descriptive tables & correlation matrices


```{r}
df_btb_wide <- BtheB %>% 
  dplyr::mutate(id = row_number()) %>% # create a new variable to ID participants
  dplyr::select(id, treatment,         # specify that ID variable is first
                drug, length,
                bdi.pre, bdi.2m, bdi.4m, bdi.6m, bdi.8m)
```

```{r}
df_btb_wide %>% 
  psych::headTail(top = 10, bottom = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Wide Format")
```

#### Long Format

Restructure to long format

**LONG FORMAT**
* One line per observation
* Good for plots and modeling

```{r}
df_btb_long <- df_btb_wide %>% 
  tidyr::pivot_longer(cols = c(bdi.2m, bdi.4m, bdi.6m, bdi.8m),
                      names_to = "month", 
                      names_pattern = "bdi.(.)m",
                      values_to = "bdi") %>% 
  dplyr::mutate(month = as.numeric(month)) %>% 
  dplyr::filter(complete.cases(id, bdi, treatment, month)) %>% 
  dplyr::arrange(id, month) %>% 
  dplyr::select(id, treatment, drug, length, bdi.pre, month, bdi)
```


```{r}
df_btb_long %>% 
  psych::headTail(top = 10, bottom = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Long Format")
```

## EXPLORATORY DATA ANLAYSIS

### Visualize: Person-profile Plots

Create spaghetti plots of the raw, observed data

```{r, fig.cap="Observed Trajectories of Depression Across Time with LM Smoother by Treatment"}
df_btb_long %>% 
  dplyr::mutate(treatment = treatment %>% 
                  forcats::fct_recode("Treatment as Usual" = "TAU",
                                      "Beat the Blues Program" = "BtheB")) %>% 
  ggplot(aes(x = month,
             y = bdi)) +
  geom_point() +
  geom_line(aes(group = id), 
            size = 1, 
            alpha = 0.3) +
  geom_smooth(method = "lm") +
  theme_bw() +
  facet_grid(.~ treatment) +
  labs(x = "Time from Baseline, months",
       y = "Beck Depression Inventory Score")
```
```{r, fig.cap="Observed Trajectories of Depression Across Time with LM Smoother by Treatment and Antidepressant Use"}
df_btb_long %>% 
  dplyr::mutate(treatment = treatment %>% 
                  forcats::fct_recode("Treatment as Usual" = "TAU",
                                      "Beat the Blues Program" = "BtheB")) %>% 
  dplyr::mutate(drug = drug %>% 
                  forcats::fct_recode("Not Using Antidepressants" = "No",
                                      "Using Antidepressants" = "Yes")) %>% 
  ggplot(aes(x = month,
             y = bdi)) +
  geom_line(aes(group = id), 
            size = 1, 
            alpha = 0.3) +
  geom_smooth(method = "lm") +
  facet_grid(drug ~ treatment) +
  theme_bw() +
  labs(x = "Time from Baseline, months",
       y = "Beck Depression Inventory Score")
```

```{r, fig.cap="Observed Trajectories of Depression Across Time with LM Smoother by Treatment, Antidepressant Use, and Length of Current Episode"}
df_btb_long %>% 
  dplyr::mutate(treatment = treatment %>% 
                  forcats::fct_recode("Treatment as Usual" = "TAU",
                                      "Beat the Blues Program" = "BtheB")) %>% 
  dplyr::mutate(drug = drug %>% 
                  forcats::fct_recode("Not Using Antidepressants" = "No",
                                      "Using Antidepressants" = "Yes")) %>% 
  ggplot(aes(x = month,
             y = bdi)) +
  geom_line(aes(group = id,
                color = length), 
            size = 1, 
            alpha = 0.3) +
  geom_smooth(aes(color = length),
              method = "lm",
              size = 1.25,
              se = FALSE) +
  facet_grid(drug ~ treatment) +
  theme_bw() +
  labs(x = "Time from Baseline, months",
       y = "Beck Depression Inventory Score",
       color = "Current Episode Length") +
  theme(legend.position = "inside",
        legend.position.inside = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.key.width = unit(2, "cm"),
        legend.background = element_rect(color = "black"))
```



```{r, fig.cap="Linear Regression Line for Subgroup Analysis by Treatment"}
df_btb_long %>% 
  dplyr::mutate(treatment = treatment %>% 
                  forcats::fct_recode("Treatment as Usual" = "TAU",
                                      "Beat the Blues Program" = "BtheB")) %>% 
  ggplot(aes(x = month,
             y = bdi,
             color = treatment,
             fill = treatment)) +
  geom_smooth(method = "lm") +
  theme_bw()  +
  labs(x = "Time from Baseline, months",
       y = "Beck Depression Inventory Score",
       color = NULL,
       fill = NULL) +
  theme(legend.position = "inside",
        legend.position.inside = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.key.width = unit(2, "cm"),
        legend.background = element_rect(color = "black"))
```

### Calculate the Observed Correlation Structure


```{r}
bdi_corr <- df_btb_wide %>% 
  dplyr::select(starts_with("bdi")) %>% 
  stats::cor(use="pairwise.complete.obs")

bdi_corr
```


```{r}
df_btb_wide %>% 
  dplyr::select(starts_with("bdi")) %>% 
  apaSupp::tab_cor(caption = "Correlation Between Repeated Measures of Depression")
```



### Plot the correlation matrix to get a better feel for the pattern

```{r}
bdi_corr %>% 
  corrplot::corrplot.mixed(upper = "ellipse")
```


## LINEAR REGRESSION, LM

This ignores any correlation between repeated measures on the same individual and treats all observations as independent.

###  Fit the models

```{r}
btb_lm_1 <- stats::lm(bdi ~ bdi.pre + length + drug + treatment + month,
                    data = df_btb_long)

btb_lm_2 <- stats::lm(bdi ~ bdi.pre + length + drug + treatment*month,
                    data = df_btb_long)

btb_lm_3 <- stats::lm(bdi ~ bdi.pre + length + drug + treatment + drug*month,
                    data = df_btb_long)

btb_lm_4 <- stats::lm(bdi ~ bdi.pre + length + drug*treatment*month,
                    data = df_btb_long)
```

### Parameter Estimates Table

```{r}
apaSupp::tab_lms(list(btb_lm_1, btb_lm_2, btb_lm_3, btb_lm_4),
                 narrow = TRUE,
                 caption = "Multiple Linear Regression - Ignoring correlation WRONG!")
```






## MULTILEVEL MODELING, MLM

### Fit the models


```{r}
btb_lmer_RI   <- lmerTest::lmer(bdi ~ bdi.pre + length + drug + treatment + 
                                  month + (1 | id), 
                                data = df_btb_long, 
                                REML = TRUE)

btb_lmer_RIAS <- lmerTest::lmer(bdi ~ bdi.pre + length + drug + treatment + 
                                  month + (month | id), 
                                data = df_btb_long, 
                                REML = TRUE,
                                control = lmerControl(optimizer = "Nelder_Mead"))
```


### Parameter Estimates Table


```{r}
apaSupp::tab_lmers(list("MLM - RI"   = btb_lmer_RI,
                        "MLM - RIAS" = btb_lmer_RIAS))
```






### Likelihood Ratio Test


```{r}
anova(btb_lmer_RI, 
      btb_lmer_RIAS, 
      refit = FALSE)
```






### Plot the model predictions

```{r, fig.cap="Estimated Marginal Means from a Multilevel Model (MLM)"}
interactions::interact_plot(model = btb_lmer_RI,
                            pred = month,
                            modx = treatment,
                            modx.labels = c("Treatment as Usual",
                                            "Beat the Blues Program"),
                            mod2 = drug,
                            mod2.labels = c("Not Using Antidepressants",
                                            "Using Antidepressants"),
                            legend.main = "Treatment",
                            interval = TRUE) +
  theme_bw()  +
  labs(x = "Time from Baseline, months",
       y = "Estimated Margial Mean\nBeck Depression Inventory Score") +
  theme(legend.position = c(0, 0),
        legend.justification = c(-0.1, -0.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```



```{r}
effects::Effect(c("treatment", "month", "drug"), 
                mod = btb_lmer_RI) %>% 
  data.frame %>%   
  dplyr::mutate(treatment = treatment %>% 
                  forcats::fct_recode("Treatment as Usual" = "TAU",
                                      "Beat the Blues Program" = "BtheB")) %>% 
  dplyr::mutate(drug = drug %>% 
                  forcats::fct_recode("Not Using Antidepressants" = "No",
                                      "Using Antidepressants" = "Yes")) %>% 
  dplyr::mutate(treatment = fct_reorder2(treatment, month, fit)) %>%
  ggplot(aes(x = month,
             y = fit)) +
  geom_line(aes(color = treatment,
                linetype = treatment),
            size = 1) +
  geom_ribbon(aes(ymin = lower,
                  ymax = upper,
                  fill = treatment), 
              alpha = 0.3) +
  theme_bw() +
  facet_grid(.~ drug)+
  labs(x = "Time from Baseline, months",
       y = "Estimated Margial Mean\nBeck Depression Inventory Score") +
  theme(legend.position = c(0, 0),
        legend.justification = c(-0.1, -0.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```




## GENERAL ESTIMATING EQUATIONS, GEE

### Explore Various Correlation Structures


#### Fit the models - Main effects to determine correlation structure


The output below each model is the 'starting' model assuming independence, so they will all be the same here.

```{r}
btb_geeglm_in <- geepack::geeglm(bdi ~ bdi.pre + length + drug + treatment + month, 
                                 data = df_btb_long, 
                                 id = id,
                                 wave = month,
                                 family = gaussian, 
                                 corstr = 'independence')

btb_geeglm_ex <- geepack::geeglm(bdi ~ bdi.pre + length + drug + treatment + month, 
                                 data = df_btb_long, 
                                 id = id, 
                                 wave = month,
                                 family = gaussian, 
                                 corstr = 'exchangeable')

btb_geeglm_ar <- geepack::geeglm(bdi ~ bdi.pre + length + drug + treatment + month, 
                                 data = df_btb_long, 
                                 id = id, 
                                 wave = month,
                                 family = gaussian, 
                                 corstr = 'ar1')

btb_geeglm_un <- geepack::geeglm(bdi ~ bdi.pre + length + drug + treatment + month, 
                                 data = df_btb_long, 
                                 id = id,
                                 wave = month,
                                 family = gaussian, 
                                 corstr = 'unstructured')
```






#### Parameter Estimates Table


```{r}
apaSupp::tab_gees(list("Exchangeable" = btb_geeglm_ex,
                       "Autoregressive" = btb_geeglm_ar,
                       "Unstructured" = btb_geeglm_un),
                  narrow = TRUE)
```




> Can't Use the Likelihood Ratio Test

The `anova()` function is used to compare nested models for parameters (fixed effects), not correlation structures.

```{r}
anova(btb_geeglm_in, btb_geeglm_ex)
anova(btb_geeglm_in, btb_geeglm_ar)
anova(btb_geeglm_in, btb_geeglm_un)
```



#### Variaous QIC Measures of Fit

References:

* Pan, W. 2001. Akaike's information criterion in generalized estimating equations. Biometrics 57:120-125. https://onlinelibrary.wiley.com/doi/abs/10.1111/j.0006-341X.2001.00120.x 

* Burnham, K. P. and D. R. Anderson. 2002. Model selection and multimodel inference: a practical information-theoretic approach. Second edition. Springer Science and Business Media, Inc., New York.
https://cds.cern.ch/record/1608735/files/9780387953649_TOC.pdf 



The `QIC()` is one way to **try** to measure model fit.  You can enter more than one model into a single function call.

QIC(I) based on independence model  <-- suggested by Pan (Biometric, March 2001), asymptotically unbiased estimator (choose the correlation stucture that produces the smallest QIC(I), p122)

```{r}
MuMIn::QIC(btb_geeglm_in, 
           btb_geeglm_ex, 
           btb_geeglm_ar, 
           btb_geeglm_un, 
           typeR = FALSE)
```


QIC(R) is based on quasi-likelihood of a working correlation R model, can **NOT** be used to select the working correlation matrix.

```{r}
MuMIn::QIC(btb_geeglm_in, 
           btb_geeglm_ex, 
           btb_geeglm_ar, 
           btb_geeglm_un, 
           typeR = TRUE)    # NOT the default
```


QIC_U(R) approximates QIC(R), and while both are useful for variable selection, they can NOT be applied to select the working correlation matrix.

```{r}
MuMIn::QICu(btb_geeglm_in, 
            btb_geeglm_ex, 
            btb_geeglm_ar, 
            btb_geeglm_un) 
```





```{r}
MuMIn::model.sel(btb_geeglm_in, 
                 btb_geeglm_ex, 
                 btb_geeglm_ar, 
                 btb_geeglm_un, 
                 rank = "QIC")     #sorts the best to the TOP, uses QIC(I)
```




### Final "Best" Model

#### Tabulate Model Parameters

```{r}
apaSupp::tab_gee(btb_geeglm_ex,
                 var_labels = c("bdi.pre" = "Baseline BDI",
                                "length" = "Current Episode Length",
                                "drug" = "Antidepressant Use",
                                "treatment" = "Intervention",
                                "month" = "Time, Months"),
                 caption = "Final GEE, exchangeable correlation")
```


#### Plot the Marginal Effects


```{r}
interactions::interact_plot(model = btb_geeglm_ex,
                            pred = "month",
                            modx = "treatment")
```







Do not worry about confidence intervals.

```{r}
expand.grid(bdi.pre = 23,
            length = "<6m",
            drug = "No",
            treatment = levels(df_btb_long$treatment),
            month = seq(from = 2, to = 8, by = 2)) %>%  
  dplyr::mutate(fit_in = predict(btb_geeglm_in,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_ex = predict(btb_geeglm_ex,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_ar = predict(btb_geeglm_ar,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_un = predict(btb_geeglm_un,
                                 newdata = .,
                                 type = "response")) %>% 
  tidyr::pivot_longer(cols = starts_with("fit_"),
                      names_to = "covR",
                      names_pattern = "fit_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(covR = factor(covR, 
                              levels = c("un", "ar", "ex", "in"),
                              labels = c("Unstructured",
                                         "Auto-Regressive",
                                         "Compound Symetry",
                                         "Independence"))) %>% 
  ggplot(aes(x = month,
             y = fit,
             linetype = treatment)) +
  geom_line(alpha = 0.6) +
  theme_bw() +
  labs(title    = "BtheB - Predictions from four GEE models (geeglm)",
       x = "Month",
       y = "Predicted BDI",
       linetype = "Treatment:") +
  scale_linetype_manual(values = c("solid", "longdash")) +
  theme(legend.key.width = unit(1, "cm")) +
  facet_wrap(~ covR)
```






```{r}
expand.grid(bdi.pre = 23,
            length = "<6m",
            drug = "No",
            treatment = levels(df_btb_long$treatment),
            month = seq(from = 2, to = 8, by = 2)) %>%  
  dplyr::mutate(fit_in = predict(btb_geeglm_in,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_ex = predict(btb_geeglm_ex,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_ar = predict(btb_geeglm_ar,
                                 newdata = .,
                                 type = "response")) %>% 
  dplyr::mutate(fit_un = predict(btb_geeglm_un,
                                 newdata = .,
                                 type = "response")) %>% 
  tidyr::pivot_longer(cols = starts_with("fit_"),
                      names_to = "covR",
                      names_pattern = "fit_(.*)",
                      values_to = "fit") %>% 
  dplyr::mutate(covR = factor(covR, 
                              levels = c("un", "ar", "ex", "in"),
                              labels = c("Unstructured",
                                         "Auto-Regressive",
                                         "Compound Symetry",
                                         "Independence"))) %>% 
  ggplot(aes(x = month,
             y = fit,
             color = covR,
             linetype = treatment)) +
  geom_line(alpha = 0.6) +
  theme_bw() +
  labs(title    = "BtheB - Predictions from four GEE models (geeglm)",
       x = "Month",
       y = "Predicted BDI",
       color = "Covariance Pattern:",
       linetype = "Treatment:") +
  scale_linetype_manual(values = c("solid", "longdash")) +
  scale_size_manual(values     = c(2, 1, 1, 1)) +
  scale_color_manual(values    = c("red", 
                                   "dodgerblue",
                                   "blue",
                                   "darkgreen")) +
  theme(legend.key.width = unit(1, "cm"))
```

### Investigate interactions NOT with time (`month`)

```{r}
btb_geeglm_ex_1 <- geepack::geeglm(bdi ~ bdi.pre*length + drug + treatment + month, 
                                   data = df_btb_long, 
                                   id = id, 
                                   wave = month,
                                   family = gaussian, 
                                   corstr = 'exchangeable')

btb_geeglm_ex_2 <- geepack::geeglm(bdi ~ bdi.pre*drug + length + treatment + month, 
                                   data = df_btb_long, 
                                   id = id, 
                                   wave = month,
                                   family = gaussian, 
                                   corstr = 'exchangeable')

btb_geeglm_ex_3 <- geepack::geeglm(bdi ~  bdi.pre*treatment + length + drug + month, 
                                   data = df_btb_long, 
                                   id = id, 
                                   wave = month,
                                   family = gaussian, 
                                   corstr = 'exchangeable')
```



```{r}
apaSupp::tab_gees(list("Episode Length" = btb_geeglm_ex_1,
                       "Antidepressant" = btb_geeglm_ex_2,
                       "Intervention" = btb_geeglm_ex_3),
                  narrow = TRUE,
                  caption = "GEE (exchangable): Interactions, not with Time")
```


### Investigate interactions with time (`month`)

```{r}
btb_geeglm_ex_11 <- geepack::geeglm(bdi ~ bdi.pre + length + drug + treatment*month, 
                                    data = df_btb_long, 
                                    id = id, 
                                    wave = month,
                                    family = gaussian, 
                                    corstr = 'exchangeable')

btb_geeglm_ex_12 <- geepack::geeglm(bdi ~ bdi.pre + length + treatment + drug*month, 
                                    data = df_btb_long, 
                                    id = id, 
                                    wave = month,
                                    family = gaussian, 
                                    corstr = 'exchangeable')

btb_geeglm_ex_13 <- geepack::geeglm(bdi ~ bdi.pre + drug + treatment +  length*month, 
                                    data = df_btb_long, 
                                    id = id, 
                                    wave = month,
                                    family = gaussian, 
                                    corstr = 'exchangeable')

btb_geeglm_ex_14 <- geepack::geeglm(bdi ~ length + drug + treatment + bdi.pre*month, 
                                    data = df_btb_long, 
                                    id = id, 
                                    wave = month,
                                    family = gaussian, 
                                    corstr = 'exchangeable')
```


```{r}
apaSupp::tab_gees(list("Intervention"   = btb_geeglm_ex_11,
                       "Antidepressant" = btb_geeglm_ex_12,
                       "Episode length" = btb_geeglm_ex_13,
                       "Baseline BDI"   = btb_geeglm_ex_14),
                  narrow = TRUE,
                  caption = "GEE (exchangable): Interactions with Time")
```



## Plot Best Model

Now only plot the significant variables for the 'best' model.


```{r}
interactions::interact_plot(model = btb_geeglm_ex,
                            pred = "month",
                            modx = "bdi.pre")
```


```{r}
interactions::interact_plot(model = btb_geeglm_ex,
                            pred = "month",
                            modx = "bdi.pre",
                            modx.values = c(10, 20, 30),
                            at = list(length = "<6m",
                                      drug = "No",
                                      treatment = "TAU"),
                            colors = rep("black", times = 3),
                            x.label = "Month",
                            y.label = "Predicted BDI",
                            legend.main    = "Baseline BDI:") +
  theme_bw() +
  theme(legend.position = "bottom",
        legend.key.width = unit(2, "cm")) +
  labs(title    = "BtheB - Predictions from the GEE model (exchangable)",
       subtitle = "Trajectory for a person with BL depression < 6 months and randomized to TAU") 
```


## Conclusion

**The Research Question**

Does the Beat-the-Blues program 

Net of other factors (use of antidepressants and length of the current episode), does the Beat-the-Blues program results in better depression trajectories over treatment as usual?

**The Conclusion**

There is no evidence that depression trajectories differ between participants randomized to the Beat the Blues program or the Treatment as Usual condition, after accounting for covariates and the correlation between repeated measurements.
