# GEE, Count Outcome: Antibiotics for Leprosy

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```




## PREPARATION 

### Load Packages


```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("sarbearschwartz/apaSupp") # updated 10/14/2025
```

Load/activate these packages

```{r, message=FALSE, error=FALSE, warning=FALSE, results='asis'}
library(tidyverse)    
library(flextable)
library(apaSupp)      # Not on CRAN, on GitHub (see above)
library(psych) 
library(lme4)         
library(lmerTest)
library(corrplot)      
library(geepack)       
library(MuMIn)  
library(interactions)
library(performance)      

library(HSAUR)        # package with the dataset
```



### Background



```{block type='rmdlink', echo=TRUE}
The following example is presented in the textbook: "Applied Longitudinal Analysis" by Garrett Fitzmaurice, Nan Laird & James Ware

The dataset maybe downloaded from: https://content.sph.harvard.edu/fitzmaur/ala/
```

> Data on count of leprosy bacilli pre- and post-treatment from a clinical trial of antibiotics for leprosy.


Source: Table 14.2.1 (page 422) in Snedecor, G.W. and Cochran, W.G. (1967).  Statistical Methods, (6th edn). Ames, Iowa: Iowa State University Press 

With permission of Iowa State University Press.

Reference: Snedecor, G.W. and Cochran, W.G. (1967). Statistical Methods, (6th edn). 
Ames, Iowa: Iowa State University Press


**The Background**

The dataset consists of count data from a placebo-controlled clinical trial of 30 patients with leprosy at the Eversley Childs Sanitorium in the Philippines. Participants in the study were randomized to either of two **antibiotics** *(denoted treatment drug A and B)* or to a **placebo** *(denoted treatment  drug C)*. 

Prior to receiving treatment, baseline data on the number of leprosy bacilli at six sites of the body where the bacilli tend to congregate were recorded for each patient. After several months of treatment, the number of leprosy bacilli at six sites of the body were recorded a second time. The outcome variable is the total count of the number of leprosy bacilli at the six sites.



**The Research Question**

In this study, the question of main scientific interest is whether treatment with antibiotics (drugs A and B) reduces the abundance of leprosy bacilli at the six sites of the body when compared to placebo (drug C). 



**The Data**

* Outcome or dependent variable(s) 

    + `count.pre` Pre-Treatment Bacilli Count
    + `count.post` Post-Treatment Bacilli Count

* Main predictor or independent variable of interest   

    + `drug` the treatment group: antibiotics (drugs A and B) or placebo (drug C)


### Enter the data by hand!

```{r}
df_raw <- tibble::tribble(
  ~drug, ~count_pre, ~count_post,
  "A",      11,     6,    "B",       6,     0,    "C",      16,    13,
  "A",       8,     0,    "B",       6,     2,    "C",      13,    10,
  "A",       5,     2,    "B",       7,     3,    "C",      11,    18,
  "A",      14,     8,    "B",       8,     1,    "C",       9,     5,
  "A",      19,    11,    "B",      18,    18,    "C",      21,    23,
  "A",       6,     4,    "B",       8,     4,    "C",      16,    12,
  "A",      10,    13,    "B",      19,    14,    "C",      12,     5,
  "A",       6,     1,    "B",       8,     9,    "C",      12,    16,
  "A",      11,     8,    "B",       5,     1,    "C",       7,     1,
  "A",       3,     0,    "B",      15,     9,    "C",      12,    20)
```


### Wide Format


```{r}
df_wide <- df_raw %>% 
  dplyr::mutate(drug = factor(drug)) %>% 
  dplyr::mutate(id = row_number()) %>% 
  dplyr::select(id, drug, count_pre, count_post)
```


```{r}
str(df_wide)
```


```{r}
df_wide %>%
  psych::headTail() %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Wide format of the Data")
```

### Long Format

```{r}
df_long <- df_wide %>% 
  tidyr::gather(key = obs,
                value = count,
                starts_with("count")) %>% 
  dplyr::mutate(time = case_when(obs == "count_pre" ~ 0,
                                 obs == "count_post" ~ 1)) %>% 
  dplyr::select(id, drug, time, count) %>% 
  dplyr::arrange(id, time)
```


```{r}
str(df_long)
```


```{r}
df_long %>%
  psych::headTail() %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Long format of the Data")
```

## Exploratory Data Analysis


### Summary Statistics



```{r}
df_long %>% 
  dplyr::group_by(drug, time) %>% 
  dplyr::summarise(N = n(),
                   M = mean(count),
                   VAR = var(count),
                   SD = sd(count)) %>% 
  flextable::as_grouped_data(groups = "drug") %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Summary for each group at each time")
```

### Visualize

```{r}
df_long %>% 
  dplyr::mutate(time_name = case_when(time == 0 ~ "Pre",
                                      time == 1 ~ "Post") %>% 
                  factor(levels = c("Pre", "Post"))) %>% 
  dplyr::mutate(drug_name = fct_recode(drug,
                                       "Antibiotic A" = "A",
                                       "Antibiotic B" = "B",
                                       "Placebo" = "C")) %>% 
  ggplot(aes(x = time_name,
             y = count)) +
  geom_line(aes(group = id)) +
  facet_grid(.~ drug_name) +
  theme_bw() +
  labs(x = NULL,
       y = "Number of Leprosy Bacilli at Six Sites of the Body")
```


```{r}
df_long %>% 
  dplyr::mutate(time_name = case_when(time == 0 ~ "Pre",
                                      time == 1 ~ "Post") %>% 
                  factor(levels = c("Pre", "Post"))) %>% 
  dplyr::mutate(drug_name = fct_recode(drug,
                                       "Antibiotic A" = "A",
                                       "Antibiotic B" = "B",
                                       "Placebo" = "C")) %>% 
  ggplot(aes(x = time,
             y = count)) +
  geom_line(aes(group = id),
            color = "gray") +
  geom_smooth(aes(group = drug),
              method = "lm") +
  facet_grid(.~ drug_name) +
  theme_bw() +
  labs(x = NULL,
       y = "Number of Leprosy Bacilli at Six Sites of the Body")
```



```{r}
df_long %>% 
  dplyr::mutate(time_name = case_when(time == 0 ~ "Pre",
                                      time == 1 ~ "Post") %>% 
                  factor(levels = c("Pre", "Post"))) %>% 
  dplyr::mutate(drug_name = fct_recode(drug,
                                       "Antibiotic A" = "A",
                                       "Antibiotic B" = "B",
                                       "Placebo" = "C")) %>% 
  ggplot(aes(x = time,
             y = count)) +
  geom_smooth(aes(group = drug,
                  color = drug_name,
                  fill = drug_name),
              method = "lm",
              alpha = .2) +
  theme_bw() +
  labs(x = NULL,
       y = "Number of Leprosy Bacilli at Six Sites of the Body",
       color = NULL,
       fill = NULL) +
  scale_x_continuous(breaks = 0:1,
                     labels = c("Pre-Treatment", "Post-Treatment"))
```



## Generalized Estimating Equations (GEE)

### Explore Various Correlation Structures


#### Fit the models - to determine correlation structure

The `gee()` function in the `gee` package 

```{r}
mod_geeglm_ind <- geepack::geeglm(count ~ drug*time,
                                  data = df_long,
                                  family = poisson(link = "log"),
                                  id = id,
                                  waves = time,
                                  corstr = "independence")

mod_geeglm_exc <- geepack::geeglm(count ~ drug*time,
                                  data = df_long,
                                  family = poisson(link = "log"),
                                  id = id,
                                  waves = time,
                                  corstr = "exchangeable")
```


```{r}
summary(mod_geeglm_ind)
```




```{r}
summary(mod_geeglm_exc)
```


#### Parameter Estimates

```{r}
apaSupp::tab_gees(list("Independence" = mod_geeglm_ind,
                       "Exchangeable" = mod_geeglm_exc),
                  caption = "GEE - Estimates on Count Scale (RR)")
```



#### Manual Extraction

```{r}
mod_geeglm_exc %>% coef() %>% exp()
```

```{r}
0.5699*1.6731
```


### Interpretation

* **Antibiotic A Group:** Starts with mean of 9.3 and drops by 45% (nearly cut in half) over the course of treatment.

* **Antibiotic B Group:** Starts at about the same mean at Antibiotic A group and experiences the same decrease.

* **Control Group (C):** Starts at about the same mean at Antibiotic A group BUT experiences a less than a 5% decrease over the student period while on the placebo pills.




### Visualize the Final Model

#### Quick

```{r}
interactions::interact_plot(model = mod_geeglm_exc,
                            pred = time,
                            modx = drug)
```



#### More Polished

```{r}
mod_geeglm_exc %>% 
  emmeans::emmeans(~ drug*time,
                   type = "response")
```


```{r}
mod_geeglm_exc %>% 
  emmeans::emmeans(~ drug*time,
                   type = "response") %>% 
  data.frame()
```




```{r}
mod_geeglm_exc %>% 
  emmeans::emmeans(~ drug*time,
                   type = "response",
                   level = .68) %>% 
  data.frame() %>% 
  dplyr::mutate(time_name = case_when(time == 0 ~ "Pre",
                                      time == 1 ~ "Post") %>% 
                  factor(levels = c("Pre", "Post"))) %>% 
  dplyr::mutate(drug_name = fct_recode(drug,
                                       "Antibiotic A" = "A",
                                       "Antibiotic B" = "B",
                                       "Placebo" = "C")) %>% 
  ggplot(aes(x = time_name,
             y = rate,
             group = drug_name %>% fct_rev,
             color = drug_name %>% fct_rev)) +
  geom_errorbar(aes(ymin = lower.CL,
                    ymax = upper.CL),
                width = .3,
                position = position_dodge(width = .25)) +
  geom_point(position = position_dodge(width = .25)) +
  geom_line(position = position_dodge(width = .25)) +
  theme_bw() +
  labs(x = NULL,
       y = "Estimated Marginal Mean\nNumber of Leprosy Bacilli at Six Sites of the Body",
       color = NULL) +
  theme(legend.position = c(0, 0),
        legend.justification = c(-0.1, -0.1),
        legend.background = element_rect(color = "black"))
```



## Follow-up Analysis

### Collapse the Predictor

```{r}
df_remodel <- df_long %>% 
  dplyr::mutate(antibiotic = drug %>% 
                  forcats::fct_collapse(yes = c("A", "B"),
                                        no = c("C")))
```

### Reduce the Model - `gee::gee()`

```{r}
mod_geeglm_exc2 <- geepack::geeglm(count ~ antibiotic:time ,
                                data = df_remodel,
                                family = poisson(link = "log"),
                                id = id,
                                waves = time,
                                corstr = "exchangeable")

summary(mod_geeglm_exc2)
```

### Compare Parameters


```{r}
apaSupp::tab_gees(list("Original" = mod_geeglm_exc,
                       "Refit" = mod_geeglm_exc2))
```



### Interpretation
 
* The grand mean is a count of 10.73 at pre-treatment. 

* The mean count dropped by about 40% among those on antibiotics, but there was no decrease for those on placebo pills, exp(b) = 0.592, p < .05, 95% CI [0.476, 0.74].



### Visualize


#### Quick

```{r}
interactions::interact_plot(model = mod_geeglm_exc2,
                            pred = time,
                            modx = antibiotic) +
  theme_bw()
```



#### More Polished



```{r}
mod_geeglm_exc2 %>% 
  emmeans::emmeans(~ antibiotic*time,
                   type = "response",
                   level = .68) %>% 
  data.frame() %>% 
  dplyr::mutate(time_name = case_when(time == 0 ~ "Pre",
                                      time == 1 ~ "Post") %>% 
                  factor(levels = c("Pre", "Post"))) %>% 
  ggplot(aes(x = time_name,
             y = rate,
             group = antibiotic,
             color = antibiotic)) +
  geom_errorbar(aes(ymin = lower.CL,
                    ymax = upper.CL),
                width = .3,
                position = position_dodge(width = .25)) +
  geom_point(position = position_dodge(width = .25)) +
  geom_line(position = position_dodge(width = .25)) +
  theme_bw() +
  labs(x = NULL,
       y = "Estimated Marginal Mean\nNumber of Leprosy Bacilli at Six Sites of the Body",
       color = "Antibiotic A or B") +
  theme(legend.position = c(0, 0),
        legend.justification = c(-0.1, -0.1),
        legend.background = element_rect(color = "black"))
```


```{r}
mod_geeglm_exc2 %>% 
  emmeans::emmeans(pairwise ~ time | antibiotic,
                   adjust = "none")
```

```{r}
mod_geeglm_exc2 %>% 
  emmeans::emmeans(pairwise ~ time | antibiotic,
                   type = "response",
                   adjust = "none")
```




## Conclusion

**The Research Question**

In this study, the question of main scientific interest is whether treatment with antibiotics (drugs A and B) reduces the abundance of leprosy bacilli at the six sites of the body when compared to placebo (drug C). 

**The Conclusion**

Both of these antibiotics significantly reduce leprosy bacilli from the pre-level (M = 10.7, equivalent groups at baseline) to lower (M = 6.3), compared to no change seen when on the placebo.

