# GLMM, Binary Outcome: Thailand Education (Hox ch 6)


Article: https://www.nature.com/articles/s41598-023-34478-0

https://stats.oarc.ucla.edu/other/examples/ma-hox/


chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://www.lesahoffman.com/CLDP945/CLDP945_Lecture07_Generalized.pdf



https://stats.oarc.ucla.edu/r/dae/mixed-effects-logistic-regression/


```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", 
                      fig.width   = 6,        
                      fig.height  = 4)        
```


## PREPARATION 

### Load Packages


```{r, message=FALSE, error=FALSE}
library(tidyverse)  
library(haven)
library(psych)        

library(lme4) 
library(optimx)

library(emmeans)     
library(performance)
library(interactions) 
```





### Background

NESTING
* `schoolid` identify school each student attends


OUCOME
* `re` student repeated a grade during primary school: Yes or No


PREDICTORS
* `sex`  sex: Girl or Boy
* `pped` attended preschool: Yes or No
* `msesc` school's mean SES, missing for some schools, grand mean centered



### Read in the data

```{r}
df_raw <- haven::read_sav("data/thaieduc.sav") %>% 
  haven::as_factor() %>% 
  haven::zap_label() %>% 
  haven::zap_formats() %>% 
  haven::zap_widths()
```

```{r}
tibble::glimpse(df_raw)
```


### Long Format

One line per child

```{r}
df_long <- df_raw %>%   
  dplyr::rename_all(tolower) %>% 
  dplyr::rename(re = `repeat`) %>% 
  dplyr::mutate_if(is.factor, 
                   ~forcats::fct_relabel(.x, stringr::str_to_title)) %>% 
  dplyr::mutate(re01 = case_when(re == "No" ~ 0,
                                 re == "Yes" ~ 1))
```

```{r}
tibble::glimpse(df_long)
```



```{r}
df_long %>% 
  psych::headTail(top = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Thailand Education Data on Repeating a Grade, Long Format")
```


### Wide Format

One Line Per School

```{r}
df_wide <- df_long %>% 
  dplyr::group_by(schoolid, msesc) %>% 
  dplyr::summarise(n_std = n(),
                   per_girl = 100*mean(sex == "Girl"),
                   per_pped = 100*mean(pped == "Yes"),
                   per_re   = 100*mean(re == "Yes")) %>% 
  dplyr::ungroup()
```

```{r}
tibble::glimpse(df_wide)
```



```{r}
df_wide %>% 
  psych::headTail(top = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Thailand Education Data on Repeating a Grade, Wide Format")
```

### Modeling Subset

```{r}
df_model <- df_long %>% 
  dplyr::filter(complete.cases(msesc, sex, pped, re))
```




## EXPLORATORY DATA ANALYSIS

### Sample Size

n = 7516 students

```{r}
nrow(df_model)
```

n = 411 - 55 = 356 schools

```{r}
nrow(df_wide)
```


```{r}
df_wide %>% 
  dplyr::filter(complete.cases(msesc)) %>% 
  nrow()
```


### Missing Data

```{r}
df_wide %>% 
  VIM::aggr(numbers = TRUE,   # shows the number to the far right
            prop = FALSE)     # shows counts instead of proportions
```




```{r}
df_wide %>% 
  dplyr::mutate(miss = ifelse(!is.na(msesc), "Include", "Exclude") %>% 
                  factor) %>% 
  dplyr::select(miss,
                "# Students" = n_std,
                "% Girls" = per_girl,
                "% Preschool" = per_pped,
                "% Repeat Grade" = per_re) %>% 
  apaSupp::tab1(split = "miss",
                caption = "Compare School Excluded due to Missing SES")
```


### Summarize

#### Demographics and Baseline

```{r}
df_wide %>% 
  dplyr::filter(complete.cases(msesc)) %>% 
  dplyr::select("# Students" = n_std,
                "% Girls" = per_girl,
                "% Preschool" = per_pped,
                "% Repeat Grade" = per_re) %>% 
  apaSupp::tab_desc(caption = "Description of Schools")
```


```{r}
df_wide %>% 
  dplyr::filter(complete.cases(msesc)) %>% 
  ggplot(aes(x = per_pped,
             y = per_re)) +
  geom_point() +
  theme_bw() 
```

```{r}
df_model %>% 
  dplyr::mutate(re = forcats::fct_recode(re,
                                         "Never Repeated" = "No",
                                         "Repeated a Grade" = "Yes")) %>% 
  dplyr::select(re,
                "Sex" = sex,
                "Attended Preschool" = pped,
                "SES, school mean" = msesc) %>% 
  apaSupp::tab1(split = "re",
                caption = "Compare Students ",
                type = list("Attended Preschool" = "categorical"))
```



## MULILEVEL ANALYSIS


### Fit Model

> Great articles:  
> *https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html#what-methods-are-available-to-fit-estimate-glmms
> * https://www.ellessenne.xyz/blog/2022/06/18/
> * https://bbolker.github.io/goettingen_2019/notes/glmm_details.html 
> slides: https://people.math.aau.dk/~rw/Undervisning/TostaII/Slides/2.pdf



#### Default is the Laplace ML

> `nAGQ = 1` (Laplace Approximation): This is the default method for glmer(). It approximates the integrand (the likelihood function) by a normal distribution using a second-order Taylor expansion around the mode. **It is more accurate than PQL but slower.**


```{r}
fit_glmer_1 <- lme4::glmer(re01 ~ sex + pped + msesc + (1|schoolid),
                           data = df_model,
                           family = binomial)
```

```{r}
summary(fit_glmer_1)
```

#### Penalized Iteratively Reweighted Least Squares

> `nAGQ = 0` (Penalized Iteratively Reweighted Least Squares - PIRLS): This is a **faster but less exact** method that avoids explicit integration by optimizing the fixed and random effects simultaneously in a penalized fashion. It is often referred to as penalized quasi-likelihood (PQL), and provides a good starting point for more accurate methods.

```{r}
fit_glmer_0 <- lme4::glmer(re01 ~ sex + pped + msesc + (1|schoolid),
                           data = df_model,
                           family = binomial,
                           nAGQ = 0)
```

```{r}
summary(fit_glmer_0)
```

#### Adaptive Gauss-Hermite Quadrature

> `nAGQ = k` (Adaptive Gauss-Hermite Quadrature - AGHQ): For `k > 1`, glmer() uses adaptive Gauss-Hermite quadrature, which is the **most reliable and accurate** method among the available options. This method approximates the integral as a weighted sum over a specific number of points (nodes).Larger values of k (e.g., up to 25) increase accuracy but also significantly increase computation time. Limitation: AGHQ in glmer() is currently implemented only for models with a single scalar random effect (i.e., **a single random effects** term with a single variance component). 

```{r}
fit_glmer_50 <- lme4::glmer(re01 ~ sex + pped + msesc + (1|schoolid),
                            data = df_model,
                            family = binomial,
                            nAGQ = 50)
```

```{r}
summary(fit_glmer_50)
```




### Parameter Estiamtes Table

```{r}
apaSupp::tab_lmers(list("PQL"     = fit_glmer_0,
                        "Laplase" = fit_glmer_1,
                        "Numeric" = fit_glmer_50),
                   narrow = TRUE,
                   var_labels = c("sex" = "Sex",
                                  "pped" = "Attended Preschool",
                                  "msesc" = "School SES"),
                   general_note = "School SES is grand mean centered.")
```




### Predicted Probabilites


#### Likert Scale

```{r}
fit_glmer_1 %>% 
  emmeans::emmeans(~ pped + sex)
```


#### Back-Transform

```{r}
fit_glmer_1 %>% 
  emmeans::emmeans(~ pped + sex,
                   type = "response")
```


```{r}
fit_glmer_1 %>% 
  emmeans::emmeans(~ pped + sex,
                   type = "response") %>% 
  data.frame() %>% 
  dplyr::mutate(across(c(prob, asymp.LCL, asymp.UCL),
                       ~apaSupp::p_num(.x, stars = FALSE))) %>% 
  dplyr::mutate(cell = glue::glue("{prob} [{asymp.LCL}, {asymp.UCL}]")) %>% 
  dplyr::select(Sex = sex,
                "Attend Preschool" = pped, 
                "Est [95% CI]" = cell) %>% 
  flextable::as_grouped_data(groups = "Sex") %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Predicted Probability of Repeating a Primary Grade, by Sex and Preschool Attendance",
                     general_note = "Predications are for grand-mean school SES.")
```


### Plot

#### Quick

```{r}
interactions::cat_plot(model = fit_glmer_1,
                       pred = pped,
                       modx = sex,
                       geom = "line")
```



#### Publishable (95% CI)

```{r}
fit_glmer_1 %>% 
  emmeans::emmeans(~ pped + sex,
                   type = "response") %>% 
  data.frame() %>% 
  dplyr::mutate(sex = forcats::fct_rev(sex)) %>% 
  ggplot(aes(x = pped,
             y = prob,
             group = sex,
             color = sex)) +
  geom_errorbar(aes(ymin = asymp.LCL,
                    ymax = asymp.UCL),
                width = .25) +
  geom_line(aes(linetype = sex),
  ) +
  geom_point(aes(shape = sex),
             size = 3) +
  theme_bw() +
  scale_color_manual(values = c("dodgerblue", "red")) +
  scale_linetype_manual(values = c("longdash", "solid")) +
  labs(x = "Attended Preschool",
       y = "Predicted Probabiliy of\nRepeated a Primary Grade",
       color = NULL,
       linetype = NULL,
       shape = NULL) +
  theme(legend.position = "inside",
        legend.position.inside = c(1.0, 1.0),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```





