# GLMM, Proportion Outcome: Meta Analysis (Hox ch 6)


See Appendix E of the Joop Hox text.

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", 
                      fig.width   = 6,        
                      fig.height  = 4)        
```




https://uoepsy.github.io/msmr/2122/lectures/msmr_lec02_LogisticMLM.html#1

https://commresearch.arizona.edu/classes/comm640/640_Book/docs/multilevel-logistic-regression.html

https://stats.stackexchange.com/questions/544937/when-is-it-appropriate-to-set-nagq-0-in-glmer

https://bookdown.org/animestina/phd_july_19/now-for-advanced-logistic-mixed-effects.html


## PREPARATION 

### Load Packages


```{r, message=FALSE, error=FALSE}
library(tidyverse)  
library(haven)
library(psych)        

library(lme4) 
library(optimx)

library(emmeans)     
library(performance)
library(interactions) 
```





### Background

NESTING
* `source` study id on which multiple conditions were reported (most studies reported on 2 of the 3 modes)



Study Characteristics:  

* `respisrr` type of study reported proportions
+ `Cr` completion rate, # respondents / total sample size
+ `Rr` response rate,  sample size was corrected for sampling frame errors  

* `year` 0 = 1947, the oldest year
* `saliency` saliency of the questionnaire, 0 = not at all, 1 = somewhat, 2 = highly

Condition Characteristics:

* `mode` dta collection method, 3-levels: Face-to-face, Telephone, or Mail
* `tel` dummy code for mode = Telephone
* `mail` dummy code for mode = Mail

* `response` either the completion rate or response rate
* `denom` number of respondents approached



### Read in the data


```{r}
df_raw <- haven::read_sav("data/metaresp.sav") %>% 
  haven::as_factor() %>% 
  haven::zap_label() %>% 
  haven::zap_formats() %>% 
  haven::zap_widths() 
```


```{r}
tibble::glimpse(df_raw)
```



### Long Format

One line per mode per source

```{r}
df_long <- df_raw %>% 
  dplyr::rename_all(tolower) %>% 
  dplyr::mutate_if(is.factor, ~forcats::fct_relabel(.x, stringr::str_to_title)) %>% 
  dplyr::mutate(res  = round(response*denom, 0)) %>% 
  dplyr::mutate(tel  = as.numeric(mode == "Tel"))%>% 
  dplyr::mutate(mail = as.numeric(mode == "Mail")) %>% 
  dplyr::mutate(yr = year + 1947) %>% 
  dplyr::mutate(yr47 = year) %>% 
  dplyr::mutate(yr77 = year - 30) %>% 
  dplyr::select(source, 
                yr, yr47, yr77,
                saliency,
                mode, tel, mail,
                respisrr,
                response, res, denom)
```

```{r}
tibble::glimpse(df_long)
```










```{r}
df_long %>% 
  dplyr::select(source, respisrr, mode, tel, mail, yr, saliency, response, denom, res) %>% 
  psych::headTail(top = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Meta Data for Response Rate, Long Format")
```

### Wide Format

One line per source

```{r}
df_wide <- df_long %>% 
  dplyr::group_by(source) %>% 
  dplyr::mutate(n_modes = n()) %>% 
  dplyr::mutate(respisrr = dplyr::first(respisrr)) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(source, n_modes, yr, yr47, yr77, respisrr, saliency, mode, response) %>% 
  tidyr::pivot_wider(names_from = mode,
                     names_prefix = "resp_",
                     values_from = response)
```


```{r}
tibble::glimpse(df_wide)
```



```{r}
df_wide %>%  
  dplyr::select(-yr47, -yr77) %>% 
  psych::headTail(top = 10) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Meta Data for Response Rate, Wide Format")
```





## EXPLORATORY DATA ANLAYSIS

### Sample Size


n = 105 modes/sources

```{r}
nrow(df_long)
```


n = 48 sources

```{r}
nrow(df_wide)
```

### Missing Data - None!

```{r}
df_long %>% 
  VIM::aggr(numbers = TRUE,   # shows the number to the far right
            prop = FALSE)     # shows counts instead of proportions
```


### Descriptive Summary

```{r}
df_wide %>% 
  dplyr::select(n_modes,
                yr,
                resp_Ftf,
                resp_Tel,
                resp_Mail) %>% 
  apaSupp::tab_desc(caption="Descriptive Summary of Sources")
```


```{r}
df_long %>% 
  dplyr::select(mode,
                yr, saliency,
                respisrr,
                response, res, denom) %>% 
  apaSupp::tab1(split = "mode")
```







```{r, fig.cap ="Distribution of Source Year (0 for 1947)"}
df_wide %>% 
  apaSupp::spicy_histos(yr)
```



## MULTILEVEL MODELING


### Base Model

#### PQL

```{r}
fit_glmer_0 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + (1|source),
                              data = df_long,
                              family = binomial,
                              nAGQ = 0)
```


```{r}
summary(fit_glmer_0)
```

#### Laplace

```{r}
fit_glmer_1 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + (1|source),
                              data = df_long,
                              family = binomial)
```


```{r}
summary(fit_glmer_1)
```

#### Compare


```{r}
apaSupp::tab_lmers(list("PQL" = fit_glmer_0,
                        "Laplace" = fit_glmer_1))
```





### Main Effect of Model

#### Random Intercepts

```{r}
fit_glmer_2 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + mode + (1|source),
                              data = df_long,
                              family = binomial)
```

```{r}
summary(fit_glmer_2)
```

#### Add Random Slopes

```{r}
fit_glmer_3 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + mode + 
                                (1|source) + (0 + tel|source) + (0+ mail|source),
                              data = df_long,
                              family = binomial)
```


```{r}
summary(fit_glmer_3)
```

#### Compare

```{r}
apaSupp::tab_lmers(list("RI" = fit_glmer_2,
                        "RIAS" = fit_glmer_3))
```

```{r}
apaSupp::tab_lmer_fits(list("RI" = fit_glmer_2,
                            "RIAS" = fit_glmer_3))
```

```{r}
anova(fit_glmer_2, fit_glmer_3)
```





### Cross-Level Interaction

Helps for "failed to converge": https://cran.r-project.org/web/packages/lme4/vignettes/lmerperf.html 

```{r}
fit_glmer_4 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + mode + 
                                saliency + yr77 +
                                (1|source) + (0 + tel|source) + (0+ mail|source),
                              data = df_long,
                              family = binomial,
                              control=glmerControl(optimizer="bobyqa",
                                                   optCtrl=list(maxfun=100000)))
```

```{r}
summary(fit_glmer_4)
```


```{r}
fit_glmer_5 <- lme4::glmer(cbind(res, denom - res) ~ respisrr + mode + 
                                saliency + mode*yr77 +
                                (1|source) + (0 + tel|source) + (0+ mail|source),
                              data = df_long,
                              family = binomial,
                              control=glmerControl(optimizer="bobyqa",
                                                   optCtrl=list(maxfun=100000)))
```


```{r}
summary(fit_glmer_5)
```

#### Compare

```{r}
apaSupp::tab_lmers(list("Main Effect" = fit_glmer_4,
                        "Intraction" = fit_glmer_5))
```


```{r}
apaSupp::tab_lmer_fits(list("Main Effect" = fit_glmer_4,
                            "Intraction" = fit_glmer_5))
```




```{r}
anova(fit_glmer_4, fit_glmer_5)
```





### Predicted Probabilities

#### At the Mean Year (1977)

```{r}
fit_glmer_5 %>% 
  emmeans::emmeans(~ mode*saliency,
                   at = list(yr77 = 0,
                             saliency = c("Not", "High")),
                   type = "response")
```


#### Each Mode Across Years

```{r}
fit_glmer_5 %>% 
  emmeans::emmeans(~ mode*yr77,
                   at = list(yr77 = c(1950-1977, 
                                      1977-1977, 
                                      1990-1977)),
                   type = "response")
```



```{r}
fit_glmer_5 %>% 
  emmeans::emmeans(~ mode*yr77,
                   at = list(yr77 = c(1950-1977, 
                                      1977-1977, 
                                      1990-1977)),
                   type = "response") %>% 
  data.frame() %>% 
  dplyr::mutate(year = as.character(1977 + yr77)) %>% 
  dplyr::mutate(mode = mode %>% 
                  forcats::fct_recode("Face-to-Face" = "Ftf",
                                      "Telephone" = "Tel",
                                      "Mail" = "Mail")) %>% 
  dplyr::mutate(across(c(prob, asymp.LCL, asymp.UCL),
                       ~apaSupp::p_num(.x, stars = FALSE))) %>% 
  dplyr::mutate(cell = glue::glue("{prob} [{asymp.LCL}, {asymp.UCL}]")) %>% 
  dplyr::select(mode, year, cell) %>% 
  tidyr::pivot_wider(names_from = year,
                     values_from = cell) %>% 
  flextable::flextable() %>% 
  apaSupp::theme_apa(caption = "Predicted Response Rates for Each Mode at Key Years, [95% CI]",
                     general_note = "Predications are for 'some' saliency.")
```



#### Pairwise Comparisons

```{r}
fit_glmer_5 %>% 
  emmeans::emmeans(~ mode*saliency|yr77,
                   at = list(saliency = "Some",
                             yr77 = c(1950-1977, 
                                      1977-1977, 
                                      1990-1977)),
                   type = "response") %>% 
  pairs(adjust = "none")
```


```{r}
fit_glmer_5 %>% 
  emmeans::emmeans(~ mode*saliency|yr77,
                   at = list(saliency = "Some",
                             yr77 = c(1950-1977, 
                                      1977-1977, 
                                      1990-1977)),
                   type = "response") %>% 
  pairs(adjust = "none") %>% 
  data.frame() %>% 
  dplyr::mutate(yr77 = as.numeric(as.character(yr77))) %>% 
  dplyr::mutate(year = as.character(1977 + yr77)) %>% 
  dplyr::mutate(contrast = contrast %>% 
                  forcats::fct_recode("Face-to-Face vs. Telephone" = "Ftf Some / Tel Some",
                                      "Face-to-Face vs. Mail" = "Ftf Some / Mail Some",
                                      "Telephone vs. Mail" = "Tel Some / Mail Some")) %>% 
  dplyr::mutate(p.adj = p.adjust(p.value, method = "fdr")) %>% 
  dplyr::mutate(across(c(p.value, p.adj), apaSupp::p_num)) %>% 
  dplyr::select(Year = year, 
                "Comparing Modes" = contrast, 
                OR = odds.ratio,
                p_Unadj = p.value,
                p_Adj = p.adj) %>% 
  flextable::as_grouped_data(groups = "Year") %>% 
  flextable::flextable() %>% 
  flextable::separate_header() %>% 
  apaSupp::theme_apa(caption = "Pairwise Comparisons Between Survey Mode at Key Years",
                     general_note = "Significance is given both unadjusted (Unadj) and adjusted (Adj) via the method of Benjamini, Hochberg, and Yekutieli to control the false discovery rate (FDR).")
```



### Plot

#### Simple

```{r}
interactions::interact_plot(model = fit_glmer_5,
                            pred = yr77,
                            modx = mode,
                            interval = TRUE,
                            int.width = .685)
```

#### Publishable

```{r, fig.cap="Predicted Response Rates Over the Years (Hox figure 6.2 on page 122)"}
fit_glmer_5 %>% 
  emmeans::emmeans(~ yr77*mode,
                   at = list(yr77 = seq(from = 1947 - 1977, 
                                        to = 1992 - 1977, 
                                        by = 1)),
                   type = "response",
                   level = .685) %>% 
  data.frame() %>% 
  dplyr::mutate(year = 1977 + yr77) %>% 
  dplyr::mutate(mode = mode %>% 
                  forcats::fct_recode("Face-to-Face" = "Ftf",
                                      "Telephone" = "Tel",
                                      "Mail" = "Mail")) %>% 
  ggplot(aes(x = year,
             y = prob,
             linetype = mode)) +
  # geom_ribbon(aes(ymin = asymp.LCL,
  #                 ymax = asymp.UCL),
  #             alpha= .15) +
  geom_line(linewidth = 1) +
  labs(x = NULL,
       y = "Predicted Response Rate",
       linetype = "Collection Mode") +
  theme_bw() +
  geom_vline(xintercept = c(1950, 1977, 1990),
             linewidth = 3,
             alpha = .2) +
  scale_linetype_manual(values = c("solid", "longdash", "dotted")) +
  theme(legend.position = "inside",
        legend.position.inside = c(.92, 1.0),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```












https://glennwilliams.me/r4psych/mixed-effects-models.html#a-note-on-power-effect-sizes-and-pairwise-comparisons



