# GLMM, Count Outcome: Bolous



```{block type='rmdlink', echo=TRUE}
GLMM FAQ
by: Ben Bolker and others

https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html
```

```{r, include=FALSE}
knitr::opts_chunk$set(comment     = "",
                      echo        = TRUE, 
                      warning     = FALSE, 
                      message     = FALSE,
                      fig.align   = "center", # center all figures
                      fig.width   = 6,        # set default figure width to 4 inches
                      fig.height  = 4)        # set default figure height to 3 inches
```

## Packages

### CRAN

```{r, message=FALSE, error=FALSE}
library(tidyverse)    # all things tidy
library(pander)       # nice looking genderal tabulations
library(furniture)    # nice table1() descriptives
library(texreg)       # Convert Regression Output to LaTeX or HTML Tables
library(psych)        # contains some useful functions, like headTail

library(lme4)         # Linear, generalized linear, & nonlinear mixed models
library(gee)          # Generalized Estimating Equations
library(optimx)       # Optimizers for use in lme4::glmer
library(MuMIn)        # Multi-Model Inference (caluclate QIC)
library(effects)      # Plotting estimated marginal means
library(interactions)
library(performance)

library(patchwork)    # combining graphics\


library(cold)   # this package holds the dataset we need
```


### GitHub

Helper `extract` functions for exponentiating parameters form generalized regression models within a `texreg` table of model parameters.

```{r, message=FALSE, error=FALSE}
# install.packages("devtools")
# library(devtools)
# install_github("SarBearSchwartz/texreghelpr")

library(texreghelpr)
```


## Data Prep

> Bolus data from Weiss 2005

Patient controlled analgesia (PCA) comparing 2 dosing regimes to self-control pain 


The dataset has the number of requests per interval in 12 successive four-hourly intervals following abdominal surgery for 65 patients in a clinical trial to compare two groups (bolus/lock-out combinations).

Bolus=  Large dose of medication given (usually intravenously by direct infusion injection or gravity drip) to raise blood-level concentrations to a therapeutic level

A ‘lock-out’ period followed each dose, where subject may not administer more medication.  

* Lock-out time is twice as long in 2mg/dose group
* Allows for max of 30 dosages in 2mg/dose and 60 in 1mg/dose group in any 4-hour period
* No responses neared these upper limits



**Variable List:**

* Indicators   

+ `id` unique subject indicator
+ `time` 11 consecutive 4-hour periods: 0, 1, 2, ..., 10


* Outcome or dependent variable    

+ `count` # doses recorded for in that 4-hour period


* Main predictor or independent variable of interest   

+ `group`1mg/dose group, 2mg/dose group



### Import

```{r}
data("bolus", package = "cold")

df_raw <- bolus
```


```{r}
str(df_raw) # see the structure
```


```{r}
psych::headTail(df_raw)
```

### Long Format

```{r}
df_long <- df_raw %>% 
  dplyr::rename(count = y) %>% 
  dplyr::mutate(id = factor(id)) %>% 
  dplyr::mutate(time = time - 1) %>% 
  dplyr::mutate(timeF = factor(time)) %>% 
  dplyr::mutate(group = factor(group))
```


```{r}
str(df_long) # see the structure
```


```{r}
psych::headTail(df_long)
```

### Wide Format

```{r}
df_wide <- df_long %>% 
  dplyr::select(-timeF) %>%
  tidyr::pivot_wider(names_from = time,
                     names_prefix = "time_",
                     values_from = count)
```


```{r}
str(df_wide) # see the structure

psych::headTail(df_wide)
```


## Exploratory Data Analysis

### Summary Statistics

#### Demographics and Baseline

```{r}
df_wide %>% 
  dplyr::select(group,
                "Count at Baseline" = time_0) %>% 
  apaSupp::tab1(split = "group")
```


#### Status over Time


```{r}
df_wide %>% 
  dplyr::select(group,
                time_0, 
                time_1, time_2, time_3, time_4, time_5, 
                time_6, time_7, time_8, time_9, time_10,
                time_11) %>% 
  apaSupp::tab1(split = "group")
```


#### Correlation over Time

```{r}
df_wide %>% 
  dplyr::select(starts_with("time_")) %>% 
  cor() %>% 
  corrplot::corrplot.mixed()
```



### Visualize

#### Person Profile Plots, by Group

```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_line(aes(group = id),
            color = "black",
            alpha = .4) +
  facet_grid(. ~ group) + 
  theme_bw()
```



```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(aes(group = id),
              method = "lm",
              color = "gray",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```



```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(aes(group = id),
              method = "lm",
              formula = y ~ poly(x, 2),
              color = "gray",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```




```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_line(aes(group = id)) +
  geom_smooth(method = "loess",
              color = "blue",
              se = FALSE) +
  geom_smooth(method = "lm",
              color = "red",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```

```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = log(count + .1))) +
  geom_line(aes(group = id)) +
  geom_smooth(method = "loess",
              color = "blue",
              se = FALSE) +
  geom_smooth(method = "lm",
              color = "red",
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw()
```


```{r}
df_long %>% 
  ggplot(aes(x = time, 
             y = count)) +
  geom_smooth(aes(color = "1. Loess"),
              method = "loess",
              se = FALSE) +
  geom_smooth(aes(color = "2. Linear"),
              method = "lm",
              se = FALSE) +
  geom_smooth(aes(color = "3. Quadratic"),
              method = "lm",
              formula = y ~ poly(x, 2),
              se = FALSE) +
  geom_smooth(aes(color = "4. Cubic"),
              method = "lm",
              formula = y ~ poly(x, 3),
              se = FALSE) +
  facet_grid(. ~ group) + 
  theme_bw() +
  labs(color = "Type of Smoother") +
  theme(legend.position = "bottom",
        legend.key.width = unit(1.5, "cm"))
```

#### Means over Time, BY Group

```{r}
df_long %>% 
  dplyr::group_by(group, timeF) %>% 
  dplyr::summarise(M = mean(count)) %>% 
  ggplot(aes(x = timeF,
             y = M,
             group = group,
             color = group)) +
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Sample Mean Count",
       color = "Dose") 
```


```{r}
df_long %>% 
  dplyr::group_by(group, timeF) %>% 
  dplyr::summarise(M = mean(log(count + .1))) %>% 
  ggplot(aes(x = timeF,
             y = M,
             group = group,
             color = group)) +
  geom_point() + 
  geom_line() + 
  theme_bw() + 
  labs(x = "Four-week Intervals",
       y = "Sample Mean Count",
       color = "Dose") 
```


## GEE Analysis

### Fit Various Correlation Structures



```{r}
mod_gee_ex <- geepack::geeglm(count ~ group + time + I(time^2),
                              id = id,
                              data = df_long,
                              family = poisson(link = "log"),
                              corstr = 'exchangeable')

mod_gee_ar <- geepack::geeglm(count ~ group + time + I(time^2),
                              id = id,
                              data = df_long,
                              family = poisson(link = "log"),
                              corstr = "ar1")	

mod_gee_un <- geepack::geeglm(count ~ group + time + I(time^2),
                              id = id,
                              data = df_long,
                              family = poisson(link = "log"),
                              corstr = "unstructured")	
```

### Compare Corelation Structures

#### QIC: Model Fit 

```{r}
MuMIn::model.sel(mod_gee_ex, 
                 mod_gee_ar, 
                 mod_gee_un, 
                 rank = "QIC")    #sorts the best to the TOP, uses QIC(I) to choose corelation structure
```


```{r}
performance::compare_performance(mod_gee_ex, 
                                 mod_gee_ar, 
                                 mod_gee_un,
                                 rank = TRUE)
```


#### Model Parameter Table



```{r}
apaSupp::tab_gees(list("Exchangable"    = mod_gee_ex,
                       "AutoRegressive" = mod_gee_ar,
                       "Unstructured"   = mod_gee_un),
                  narrow = TRUE,
                  caption = "GEE: Main Effects Only, with Quadratic Time")
```

### Add Interaction Terms

```{r}
mod_gee_ex2 <- geepack::geeglm(count ~ group + time,
                               id = id,
                               data = df_long,
                               family = poisson(link = "log"),
                               corstr = "exchangeable")


mod_gee_ex3 <- geepack::geeglm(count ~ group*time + group*I(time^2),
                               id = id,
                               data = df_long,
                               family = poisson(link = "log"),
                               corstr = "exchangeable")
```




```{r}
MuMIn::QIC(mod_gee_ex, 
           mod_gee_ex2,
           mod_gee_ex3, 
           typeR = TRUE)    # NOT the default
```



```{r}
apaSupp::tab_gees(list("Quadratic" = mod_gee_ex,
                       "Linear" = mod_gee_ex2,
                       "Interaction" = mod_gee_ex3),
                  narrow = TRUE,
                  caption = "GEE: Exchangeable")
```


### Visualize Best Model

#### Model Parameter Table

```{r}
apaSupp::tab_gee(mod_gee_ex2)
```






#### Predict over a Grid 

Estimated Marginal Mean Counts


```{r}
mod_gee_ex2 %>% 
  emmeans::emmeans(pairwise ~group|time,
                   at = list(time = c(0, 5, 10)),
                   type = "response")
```
       

#### Plot Estimated Marginal Means



```{r}
interactions::interact_plot(model = mod_gee_ex2,
                            pred = time,
                            modx = group) # Can NOT do interval ribbons
```


```{r}
mod_gee_ex2 %>% 
  emmeans::emmeans(~group|time,
                   at = list(time = seq(from = 0, to = 11, by = .1)),
                   type = "response",
                   level = .685) %>% 
  data.frame() %>% 
  ggplot(aes(x = time,
             y = rate,
             group = group)) +
  geom_ribbon(aes(ymin = lower.CL,
                  ymax = upper.CL,
                  fill = group),
              alpha = .2) +
  geom_line(aes(color = group,
                linetype = group)) + 
  theme_bw() + 
  labs(x = "Four-Hour Intervals",
       y = "Estimated Rate\nof Doses per 4 Hour Window",
       color = "Dose",
       fill = "Dose",
       linetype = "Dose") + 
  scale_x_continuous(breaks = seq(from = 0, to = 12, by = 2)) +
  theme(legend.position = "inside",
        legend.position.inside = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```



## GLMM Analysis


### RI: Random Intercepts Only

```{r}
mod_glmer_ri <- lme4::glmer(count ~ group + I(time/4) + I((time/4)^2) + (1|id),
                            data = df_long,
                            family = poisson(link = "log"))

summary(mod_glmer_ri)
```


### RIAS: Random Intercepts and Slopes

```{r}
mod_glmer_rias <- lme4::glmer(count ~ group + I(time/4) + I((time/4)^2) + 
                                (I(time/4)|id),
                              data = df_long,
                              family = poisson(link = "log"))

summary(mod_glmer_rias)
```



```{r}
anova(mod_glmer_ri, mod_glmer_rias)
```

```{r}
apaSupp::tab_glmers(list("RI"   = mod_glmer_ri, 
                        "RIAS" = mod_glmer_rias),
                    caption = "GLMM: Main Effects Models")
```




### RAIS: Add Interaction


See the [GLMM - Optimizers page](https://cehs-research.github.io/eBook_multilevel/glmm-binary-outcome-antibiotics-for-leprosy.html#glmm---optimizers) for more information on convergence problems.

```{r}
mod_glmer_rias2 <- lme4::glmer(count ~ group*I(time/4) + group*I((time/4)^2) + 
                                 ( I(time/4)|id ),
                               data = df_long,
                               control = glmerControl(optimizer ="Nelder_Mead"), 
                               family = poisson(link = "log"))	
```

```{r}
anova(mod_glmer_rias, mod_glmer_rias2)
```

```{r}
apaSupp::tab_glmers(list("Main Effects" = mod_glmer_rias,
                         "interactions" = mod_glmer_rias2),
                    caption = "GLMM: Main Effects vs. Interaction")
```

### RIAS: Simplify


```{r}
mod_glmer_rias3 <- lme4::glmer(count ~ group + I(time/4)  + 
                                 ( I(time/4)|id ),
                               data = df_long,
                               control = glmerControl(optimizer ="Nelder_Mead"), 
                               family = poisson(link = "log"))	
```

```{r}
anova(mod_glmer_rias, mod_glmer_rias3)
```

```{r}
apaSupp::tab_glmers(list("Quadratic" = mod_glmer_rias,
                         "Linear" = mod_glmer_rias3),
                    caption = "GLMM: Nature of Time")
```


### Visualize Best Model

#### Model Parameter Table

```{r}
apaSupp::tab_glmer(mod_glmer_rias3)
```



#### Estimated Marginal Means

```{r}
mod_glmer_rias3 %>% 
  emmeans::emmeans(pairwise ~group|time,
                   at = list(time = c(0, 5, 10)),
                   type = "response")
```

#### Plot Estimated Marginal Means


```{r}
interactions::interact_plot(model = mod_glmer_rias3,
                            pred = time,
                            modx = group,
                            interval = TRUE,
                            int.width = .685)
```


```{r}
mod_glmer_rias3 %>% 
  emmeans::emmeans(~group|time,
                   at = list(time = seq(from = 0, to = 11, by = .1)),
                   type = "response",
                   level = .685) %>% 
  data.frame() %>% 
  ggplot(aes(x = time,
             y = rate,
             group = group)) +
  geom_ribbon(aes(ymin = asymp.LCL,
                  ymax = asymp.UCL,
                  fill = group),
              alpha = .2) +
  geom_line(aes(color = group,
                linetype = group)) + 
  theme_bw() + 
  labs(x = "Four-Hour Intervals",
       y = "Estimated Rate\nof Doses per 4 Hour Window",
       color = "Dose",
       fill = "Dose",
       linetype = "Dose") + 
  scale_x_continuous(breaks = seq(from = 0, to = 12, by = 2)) +
  theme(legend.position = "inside",
        legend.position.inside = c(1, 1),
        legend.justification = c(1.1, 1.1),
        legend.background = element_rect(color = "black"),
        legend.key.width = unit(1.5, "cm"))
```

